---
title: "Wrangling and (more) Data Verbs"
author: 
  - Instructor - Soumya Mukherjee
  - Content Credit- Dr. Matthew Beckman and Olivia Beck
date: "July 17, 2023"
output: 
    slidy_presentation: default
    html_notebook: default
---


```{r include=FALSE}
# Frontmatter
rm(list = ls())

# packages
library(mosaicData)
library(knitr)
library(tidyverse)
library(dcData)

```


# Agenda

- Review of data verbs & summary functions
- Discussion of new functions and key concepts
- Guided example highlighting several new data verbs
- RStudio Concept Maps: <https://github.com/rstudio/concept-maps>


# RMarkdown Stuff

## R Markdown - Text versus R chunk

When we want to write plain text, we just write as normal. 

When we want to use a header we use # before our text 

When we want to write R code, we use an R chuck:

```{r}
x <- 1:10
sum(x)
```

```{r}

```






## R Notebook vs R Markdown Comparison

  |                  |  R Notebook            |  R Markdown      |
  |:-----------------|:----------------------:|:----------------:|
  | Source document  | FileName.Rmd           | FileName.Rmd     |
  | Typesetting syntax | markdown             | markdown   |
  | YAML Header      | `output: html_notebook` | `output: html_document`  |
  | compile button text | "Preview" or "Preview Notebook" | "Knit to HTML" or "Knit" |
  | compile button icon  |  notebook         | blue yarn ball  |
  | Output file default  | FileName.nb.html  | FileName.html  |
  | Code chunk handling  | Run R code first  | Runs R code when rendered |
  | .Rmd source embedded? | **YES**              | **NO**    |

## Discussion/Review Task: 

- **List 1: data verbs:**
    - `mutate()`
    - `filter()`
    - `select()`
    - `arrange()`
- **List 2: more data verbs:**
    - `head()` & `tail()`
    - `rename()`
    - `sample_n()`
    - `summarise()` & `group_by()`
- **List 3: summary functions:**
    - `glimpse()`
    - `str()`
    - `summary()`
    - `nrow()` & `ncol()`
    - `names()`
    - `colnames()`
    - `View()` 


# Three Important Concepts (Again)

1. Data can be usefully organized into tables with "cases" and "variables."  With "tidy data"
    - every row corresponds to a distinct case (e.g. a person, a car, a year, a country in a year) 
    - columns represent variables/features of the cases

#. Data graphics and "glyph-ready" data
    - each case corresponds to a "glyph" (mark) on the graph
    - each variable to a graphical attribute of that glyph such as x- or y-position, color, size, length, shape, etc. 
    - similar story applies for modeling purposes

#. When data are not yet in glyph-ready form, you can transform (i.e. wrangle) them into glyph-ready form.  
    - Such transformations are accomplished by performing one or more of a small set of basic operations on data tables
    - This is the work of "data verbs" and other functions



# Presidential Investigation 

- Let's take some of our new tools for a spin...
- The `presidents` data are available from the `tidyverse` package (actually `ggplot2` which was loaded by `tidyverse`)
- we'll add a few other presidents to the data as well

```{r echo=TRUE}
library(tidyverse)
library(dcData)
library(tibble)     # we'll use `tibble::tribble()`  to specify new rows 
# install.packages(lubridate)
library(lubridate)  # nice functions for handling dates

# data intake
data("presidential", package = "ggplot2")
help(presidential)
head(presidential)
## Add some more cases to the data set 
# you can search the help for the `tribble` function if you want to figure it out what's happening
AddPres <- 
    tibble::tribble(~name, ~start, ~end, ~party,
                     "Roosevelt", ymd("1933-03-04"), ymd("1945-04-12"), "Democratic",
                    "Truman", ymd("1945-04-12"), ymd("1953-01-20"), "Democratic",
                    "Trump", ymd("2017-01-20"), ymd("2021-01-20"), "Republican", 
                    "Biden", ymd("2021-01-20"), ymd(today()), "Democratic")
AddPres
# append the rows of our two "data frames"
Presidents <- bind_rows(presidential, AddPres)
Presidents
head(Presidents)
```


## Guided practice 

1. various summary/exploration functions to meet the Presidents
2. calculate `duration` of presidency 
3. show all records for presidents named "Bush"
4. subset the data to include only `name`, `party`, `duration`
5. which 3 presidents served shortest?
6. which 3 presidents served longest?
7. which presidents served more than 4 years, but less than 8 years?
8. investigating the power balance
   - which party has most presidents?
   - which party spent most days in power?


# Answers 

```{r}
# 1. summary functions
glimpse(Presidents)  # what's the benefit?
str(Presidents)      # ?
summary(Presidents)  # not so useful?!
#View(Presidents)     # console please!


# some wrangling 
Presidents_wrangled <- 
    Presidents %>%
    mutate(duration = end - start,                                     # 2
           name = if_else(name == "Bush" & start < ymd("2000-01-01"),  # 3
                          true = "Bush (Sr)", 
                          false = name)) %>%
    select(name, party, duration)                                      # 4
    

### some results
#2 Presidency duration
Presidents_wrangled

#3 The Bushes
Presidents_wrangled %>%
  filter(name %in% c("Bush","Bush (Sr)"))

# 6 shortest
Presidents_wrangled %>%
    arrange(duration) %>%
    head(n=3)

# 7. longest  
Presidents_wrangled %>%
    arrange(duration) %>%
    tail()

Presidents_wrangled %>%
    arrange(desc(duration)) %>%
    head()

# 8. more than 4 years, but less than 8 years

Presidents_wrangled %>%
    filter(duration > 1461, duration < 2922)


# 9. party balance

Presidents_wrangled %>%
    group_by(party) %>%
    summarise(total = n(), 
              in_power = sum(duration))
```

## Things to look through before tomorrow's class (July 18)
- Read Chapters 10 and 11 of Data Computing Ebook (No reading quiz assigned yet, but if you do read beforehand, more helpful for you)
- Start looking through the PopularNames project in the DataComputing Ebook in advance (Will be assigned as an activity some time this week). Link is <https://dtkaplan.github.io/DataComputingEbook/project-popular-names.html#project-popular-names>.