---
title: "Day 4"
author:
- 'Instructor: Soumya Mukherjee'
- 'Content credit: Dr. Matthew Beckman and Olivia Beck'
date: "July 7, 2023"
output:
  slidy_presentation: default
  html_notebook: default
---

## Packages + Tidyverse
- You only need to install a package once (on every machine you're on)
```{r}
install.packages("tidyverse")

## every time you want to use a function inside a package, you need to load the library
library(tidyverse) #now we can use the tidyverse functions! 
```
- Before we get into functions in the tidyverse, we will introduce one more special operator in R
- The tidyverse package contains the `%>%`  (pipe) operator 
- Pipe takes what ever is before the `%>%` and inputs as the first element into whatever is after the `%>%`.

## An example using the mean function
```{r}
?mean 
numbers <- c(1,5,6,8,23,45,67)
```
-The following are equivalent 
- Method 1 
```{r}
mean(numbers)
```
- Method 2
```{r}
numbers %>%
  mean()
```
- Pipe is usually used when we have a series of operations we want to compute
- Lets use the iris data set as an example 
```{r}
iris
```
- Say I want to arrange the rows in this data set by the Sepal.Length. 
- The following are equivalent
- Method 1
```{r}
arrange(iris, Sepal.Length)
```
- Method 2
```{r}
iris %>%
  arrange(Sepal.Length)
```
- But now say we only want to keep rows with Sepal.Width > 3 AND arrange the table by Sepal.Length
- The following are equivalent 
- Method 1
```{r}
temp1 <- arrange(iris, Sepal.Length)
result <- filter(temp1, Sepal.Width > 3)
result 
```
- Method 2
```{r}
iris %>%
  arrange(Sepal.Length) %>%
  filter(Sepal.Width > 3)
```
- Both methods technically work, but method 2 is must easier to read and took a lot less typing! 
- (advanced) Method 2 takes up significantly less memory on your machine than Method 1. This is one of the primary reasons we use `%>%`.