---
title: "Solution: DC Exercises Chapter 3"
author:
- "Instructor - Soumya Mukherjee"
- "Content Credit- Dr. Matthew Beckman and Olivia Beck"
date: "July 17, 2023"
output: pdf_document
---

## Assigned Exercises:
- DC Ch 3 Exercises: 3.1, 3.2, 3.3, 3.4, 3.5, 3.6

## Grading:
Grade as follows:

- [3 points] completion (attempting all exercises)
- 1 point for correct solutions for all questions except 3.3
- 2 points for correct solution of 3.3

**Please be as lenient as possible while grading**.

## Solutions:

### Problem 3.1

1. a data frame:
```{r eval=F}
Fireplace
```
2. a function:
```{r eval=F}
fireplace()
```
3. the name of a named argument:
```{r eval=F}
function(x = fireplace)
```
4. a variable:
```{r eval=F}
fireplace
```

### Problem 3.2
Flawed statement:
```{r eval=F}
Result <- %>% filter(BabyNames, name == "Prince")
```
Corrected statement:
```{r eval=F}
Result <- BabyNames %>% filter(name == "Prince")
```

- pipe operator `%>%` can’t follow an assignment operator `<-`
- data table `BabyNames` should come before the pipe rather than as an argument to the filter function

### Problem 3.3

Here are the kinds of named objects given: 
a. function name
b. data table name
c. variable name
d. name of a named argument

Identify what kind of object each of the following name is used for:

1. BabyNames - b
2. filter - a
3. name - c,d
4. == - d
5. group_by - a
6. year - c, d
7. sex - c, d
8. summarise - a
9. yearlyTotal - d
10. sum - a, d
11. count - d
12. ggplot - a
13. aes - a, d
14. x - c, d
15. y - c, d
16. geom_point - a
17. color - d
18. geom_vline - a
19. xintercept - d

### Problem 3.4

- meaning of `brainwt` variable: weight of an animal brain
- how many cases: 83
- real-world meaning of case: An animal/mammal
- levels of `vore`: {carni, herbi, insecti, omni, NA}

### Problem 3.5
a. SAME
b. SAME
c. DIFFERENT. This includes a `mean()` function rather than the `sum()` function shown in the other commands
d. SAME

### Problem 3.6
1. `date()` takes no arguments
2. `date()` returns a string **constant** of the current date and time.