---
title: "Tidyverse examples, Markdown schedule Activity"
author:
- 'Instructor: Soumya Mukherjee'
- 'Content credit: Dr. Matthew Beckman and Olivia Beck'
date: "July 10, 2023"
output:
    slidy_presentation: default
    html_notebook: default
---

## A more comprehensive introduction to tidyverse
- Load the Library
```{r}
library(tidyverse)
```
- If you find an error in the above step, install tidyverse first.
- Notice the difference between these three lines
```{r}
y <- 10
y
y <- -10
y
y < - 10
y
```

## Tidyverse takes advantage of logic operators

```{r}
#------------------- Comparisons with >, <, ==, <=, >= ---------------------
# >, < is greater than and less than 
5 > 3        # True 
7 * 2 > 100  # False 
9 < 9    #False

# >=, <= greater than or equal to, and less than or equal to
5 >= 3       # True
(7 * 2) >= 100 # False 
9 <= 9       # True 

# == is "is equal to?"
10 == 2 * 5 # True
7 == 14 / 3 # False  
TRUE == TRUE # True
FALSE == FALSE # True
TRUE == FALSE # False
```

## Be careful of precision errors 
```{r}
sqrt(2) == 1.414214 # False
pi == 3.14159265358979 # False
```
## >, <, <=, >=, == are also vectorized functions
```{r}
c(5, 6) <= c(9, 12)
c(TRUE, 5 * 8) <= c(FALSE, 40)
5 == c(5, 6)
```

## Comparing NA values
```{r}
5 == NA # = NA, not FALSE #notice the caution arrow on next to the line number
NA == NA # = NA, not TRUE
# We use is.na() to test for NA values 
is.na(5) # False
is.na(NA) # True 
is.na(NA * 5) # True
```

## And" and "Or"
```{r}
# & is the "and" operator 
TRUE & FALSE #False 
TRUE & TRUE #True
FALSE & FALSE #False 
x <- 2
y <- -10
(x < 3) & (y > -50) #True
(x <= 2) & (y < -25 ) #False


## | is the "or" operator (in math, "or" mean one, the other, or both)
TRUE | FALSE #True 
TRUE | TRUE #True
FALSE | FALSE #False 
x <- 2
y <- -10
(x < 3) | (y > -50) #True
(x <= 2) | (y < -25 ) #True

## | and & are vectorized functions 
c(TRUE, TRUE, FALSE) | c(FALSE, TRUE, FALSE)
c(TRUE, TRUE, FALSE) & c(FALSE, TRUE, FALSE)
```

## Loading in a data set in Tidyverse 
- In Tidyverse use read_filetype("path/to/file.filetype")
- To read in a tsv we use 
```{r}
?read_tsv
dat.football <- read_tsv(file = "https://raw.githubusercontent.com/ada-lovecraft/ProcessingSketches/master/Bits%20and%20Pieces/Football_Stuff/data/nfl-salaries.tsv")
```

- View the top portion of the data to get an idea of what it looks like 
```{r}
head(dat.football) #default is first 6 rows and all the columns
head(dat.football, n =10)
dat.football %>%
  slice(1:10)
```

- Get the dimensions of the data
```{r}
dim(dat.football)
```
-Get the column names of the data 
```{r}
colnames(dat.football)
dat.football %>%
  colnames()
```
- Get the row names of the data
```{r}
rownames(dat.football) #meaningless! (most times they will be)
```
- Get a summary of the data 
```{r}
summary(dat.football) # gives summary info by column
```


## Basic Functions in Tidyverse 
- Tidyverse uses "verbs" as function names to describe what it is doing to the data
- Let's look at a few of these "verbs"
    - Filter 
```{r}
dat.football %>%
  filter(Team == "Denver Broncos")
```  
    - Arrange 
```{r}  
dat.football %>%
  arrange(Salary) #lowest to highest
dat.football %>%
  arrange(desc(Salary)) #highest to lowest
```
    - Select
```{r}
dat.football %>%
  select(PlayerName, Position)
```
    - Rename
```{r}
dat.football %>%
  rename(TeamName = Team)
```
    - Mutate
```{r}
dat.football %>%
  mutate(PercentOfCap = Salary / SalaryCap * 100)
```
    - Group
```{r}
dat.football %>%
  group_by(Team) #doesn't look like it did anything??? 
```
    - Summarise
```{r}
dat.football %>%
  summarise(MeanSalary = mean(Salary))

dat.football %>%
  summarize(SdSalary = sd(Salary))

?summarise

dat.football %>%
  group_by(Team) %>%
  summarise(MeanSalary = mean(Salary), .groups = "keep" )
```

## Exploratory Analysis - Combining it all together 
- What is the highest salary? 
```{r}
max(dat.football$Salary)
```
- Which player has this salary? 
    - Method 1
```{r}
max.salary <- max(dat.football$Salary) #get the max salary
row.max.salary <- dat.football$Salary == max.salary
answer.1 <- dat.football$PlayerName[row.max.salary]
answer.1
```

    - Method 2
```{r}
answer.2 <- dat.football %>%
  filter(Salary == max(Salary) ) %>%
  select(PlayerName)
```
    - Method 3
```{r}
answer.3 <- dat.football %>%
  arrange(desc(Salary)) %>%
  slice(1) %>%
  select(PlayerName)
```

## What is the benefit of using tidyverse functions?
```{r}
library(utils)
object.size(c(max.salary, row.max.salary, answer.1))
object.size(answer.2)
object.size(answer.3)
# 944/12304 ; used only 7% of the storage space by using tidyverse
```
## Further Analysis 
- Which is the team with the highest paid roster, and what was their total pay? Which is the team with the lowest paid roster, and what was their total pay?
```{r}
Paid <- dat.football %>%
  group_by(Team)%>% 
  summarize(PaidRoster = sum(Salary)) %>% 
  arrange(desc(PaidRoster))
Paid[1, ] #highest paid 
# how many teams are in our data set> 
dim(Paid)
length(unique(dat.football$Team))
Paid[31, ] 
```

## Pivot Wider and Pivot Longer

- Pivot Wider 
```{r}
?pivot_wider
# names_from = new column names 
# value_from = values to fill in in the table
us_rent_income
us_rent_income %>%
  pivot_wider(
    names_from = variable,
    values_from = c(estimate, moe)
  )
```
  - is the above table tidy? What is each case?

- Pivot Longer 
```{r}
?pivot_longer
#name_to = new column name that will contain the old column names
#values_to = new column name that will contain the data from the original table
relig_income
relig_income %>%
  pivot_longer(!religion,  # every column but religion 
               names_to = "income", 
               values_to = "count")
```
  - Is the above table Tidy? What is a case?
```{r}
world_bank_pop
world_bank_pop %>%
  pivot_longer(!c(country, indicator),
               names_to = "year",
               values_to = "count")
```
  - Is the above table tidy? What is a case?
  
## Assignments before next lecture tomorrow (July 11)
- Activity: STAT184-Rmd-schedule (Will explain in class). Assignment link: <https://classroom.github.com/a/Vbc6zqua>
- No reading Quiz assigned for tomorrow, but please read Chapter 5 and 6 of DataComputing Ebook before class tomorrow. (<https://dtkaplan.github.io/DataComputingEbook/chap-graphics-introduction.html#chap:graphics-introduction> and <https://dtkaplan.github.io/DataComputingEbook/chap-frames-glyphs.html#chap:frames-glyphs>)

- Activity: GitHub-Practive-184, DC Exercises 2 and 3 (due date extended by 1 day, so all of them due tomorrow)